---
title: "DATA1901 Cannabis Use Disorder Project"
author: "Thomas Love, Scarlett Foulsham, Lachy Miller, Liliann Lucas, and Nusrat Samiya"
date: "University of Sydney | DATA1901 | September 2021"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    code_download: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: yes
    theme: sandstone
    highlight: tango
    code_folding: hide
---

```{css newstyles, echo=FALSE}
h1, .h1, h2, .h2, h3, .h3 { /* Add space before headings: */
    margin-top: 56px;
}
h1, h2 { /* add border to h1 and h2 */
  border-bottom: solid 1px #666;
}
h2 { /* Resize header 2: */
  font-size: 22px;
}
h3 { /* Resize header 2: */
  font-size: 18px;
}
body { /* Make main text colour black */
  color: black;
}
a { /* Link colours */
  color: blue;
}
.tocify { /* Some toc fixes*/
  width: 100% !important;
  border: none; /* remove border */
}
.tocify-header { /* fix for horrible indent in toc */
  text-indent: initial;
}

```


# Executive Summary

# Data Setup
Before we begin our data exploration, let's read in the data and load some important packages.

Let's import the dataset which we've called `study_data`.
```{r}
data = read.csv("study_data.csv")
```
Let's load `tidyverse`, `gapminder`, and `plotly`.
```{r, message=FALSE}
library(tidyverse)
library(gapminder)
library(plotly)
```


# Initial Data Analysis 
Great! Now let's begin our IDA where we'll explore the data, classify the variables, isolate a variable, and visualize the data. 


## Explore the Data
Let's use `glimpse()`.
```{r}
glimpse(data)
```
We can see that the dataset has 398 rows and 4 columns which are id, group, week, and cu_total. 
We can also see that `group` and `week` are classified as `chr`, let's make them both qualitative.

## Classify the Variables
Use the `as.factor()` function.
```{r}
data$week = as.factor(data$week)
class(data$week)
data$group = as.factor(data$group)
class(data$group)
```


## Isolate a Variable
Let's isolate cu_total and use `summary()` and `table()` to produce some basic numerical summaries.
```{r}
data$cu_total
summary(data$cu_total)
table(data$week)
```
Here we can see that the minimum number of cannabis use instances was 0 
while the maximum was 28. We can also see through `table()` that there is 
an association between the *weeks progressing* and *decreased cannabis use frequency*.



## Visualize the Data
Let's create an initial visualisation of that association with `ggplot2`.
```{r, message=FALSE}

df <- data %>%
  filter(group == "nabiximols" | group == "placebo") %>%
  select(group, week, cu_total)

#To remove NA in cu_total

rf <- df %>%
  filter(cu_total != "NA") %>%
  select(group, week, cu_total)

#To remove week00

ff <- rf %>%
  filter(week != "week00") %>%
  select(group, week, cu_total)

#Histogram Without Week 00 and NA Values
ggplot(ff, aes(x=cu_total)) + geom_histogram(aes(fill=week)) + facet_wrap(~ group) + ggtitle("Histogram Without Week 00 and NA Values")

```

# Research Questions

## Number 1
> How does the relationship between the placebo and treatment group evolve throughout the 28 day study period?

```{r, results = 'hide', message=FALSE}
#Creating different levels in the data for the group and week
data %>%
  mutate(id = factor(id),
         group = factor(group,
                        levels = c("placebo", 
                                   "nabiximols")), 
         week = factor(week,
                       levels = c("week00",
                                  "week04",
                                  "week08",
                                  "week12"))) 

#Filtering out the NA values
Data <- data %>%
  filter(cu_total != "NA") %>% 
  group_by(group, week)

#Adding mean, sd, count and se for each group and week
Level_summary <- Data %>%
  group_by(group, week) %>%
  summarise(mean = mean(cu_total, 
                        na.rm = T),
            sd = sd(cu_total,
                    na.rm = T),
            count = n()) %>%
  mutate(se = round(sd/sqrt(count),2)) %>%
  relocate(se, .before = "count") 

#Line graph of average cu_total by week for each group 
  ggplot(Level_summary, mapping = aes(x = week,
                       y = mean,
                       group = group)) +
  geom_point(mapping = aes(colour = group,
                           shape = group)) +
  geom_line(mapping = aes(colour = group)) +
  coord_cartesian(ylim = c(0,28)) +
  scale_y_continuous(breaks = seq(0,28,7)) +
  theme_light() +
  theme(legend.title = element_blank(),
        panel.grid = element_blank()) +
  labs(y = "days cannabis use in previous 28 days")
```

Therefore, each group has a differing trend through out the trial period,
with the trend in mean days of cannabis use in the previous 28 days decreasing 
for nabiximols, compared to the placebo. The mean results for the placebo group 
even starts to increase towards the end of the trial. Therefore, this suggests 
that a statistically significant difference exists between the treatments and 
that nabiximols is effective in treating cannabis addiction, however, this 
still needs to be determined.

## Number 2
> Is a linear model appropriate to analyse this data?

```{r, results='hide'}
#Creating a wide table, separating the weeks.
  WideView <- Data %>%
    pivot_wider(names_from = week,
                values_from = cu_total) 
  
#Add total days use, total observable days and completed trial variables to the data.
  WideView %<>%
    rowwise() %>%
    mutate(totalDaysUse = sum(c_across(week04:week12), 
                              na.rm = T),
           totalObsDays = (3 - sum(is.na(c_across(week04:week12)), 
                                   na.rm = T))*28,
           completedTrial = factor(ifelse(totalObsDays < 84,
                                          "didNotComplete",
                                          "completed"),
                                   levels = c("didNotComplete",
                                              "completed"))) %>% ungroup()
  
  #separating only the participants that completed the trial.
  Completed <- WideView %>%
   filter(completedTrial == "completed")
 
  #Now only 67 participants are analysed because the other 61 participants did not complete the trial.
  Completed %>%
    group_by(group) %>%
    summarise(mean = mean(totalDaysUse, 
                          na.rm = T),
              sd = sd(totalDaysUse,
                      na.rm = T), 
              count = n())
  
  #Create a linear model
    Linear1 <- lm(formula = totalDaysUse ~ group + week00,
              data = Completed)
  summary(Linear1)
  
  #Analyse regression graph of linear model
  ggplot(Linear1, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "red") +
  ggtitle("Resid vs Fitted")
  
```

Therefore, this linear model is not appropriate for this data as a 
spreading out trend occurs in the residual plot as the x-value increases.

## Number 3
> Is the relationship between the number of days of cannabis use within the trial period for the nabiximpols and the placebo group for participants that completed the trial statistically significant?

```{r}
#Box Plot
ggplot(Completed, aes(x=totalDaysUse)) + geom_histogram(colour = "red", fill = "orange", alpha = 0.3) + facet_wrap(~ group) +
  theme_linedraw() +
  theme(legend.position = "none")

#Bar Plot
ggplot(Completed, aes(x = group, y = totalDaysUse)) +
  geom_boxplot(color="red", fill="orange", alpha=0.3) +
  xlab("Group") +
  ylab("Total days of usage") + 
  ggtitle("Relationship between participants group and their total days of usage through out the trial") +
  theme_linedraw() +
  theme(legend.position = "none")

```

As evident in the histogram, the total days use for participants 
in each treatment condition is not normally distributed. Further,
a because a contemporaneous control (placebo) is used, as opposed 
to historical controls, which is unpaired to the nabiximols group. 
This suggests that a Mann-Whitney U test should be applied to determine
whether the difference in total days of cannabis use within the trial 
between the two groups is statistically significant. The box plot 
also suggests that a significant difference exists, due to the 
drastic difference in medians and spreads.

```{r}
#Use a Mann-Whitney test since the data is not normally distributed
wilcox.test(totalDaysUse ~ group, data = Completed)

#Test whether the groups were effectively randomly allocated by statistically comparing the amount of days of cannabis use of the participants in the 28 days leading up to the trial
wilcox.test(week00 ~ group, data = Completed)
```

The nabiximols group has significantly lower total days use of cannabis
during the trial then the placebo group (p = 0.03872). Therefore, nabiximols 
is effective in treating cannabis addiction. The nabiximols group does not 
have significantly higher or lower total days use of cannabis during the 
28 days leading up to the trial then the placebo group (p = 0.03872). 
This suggests that the two groups were effectively randomly allocated 
and were fair to test. 

# Limitations
People have different biologies and genetic makeup, using a historical 
control would have given better understanding of what individual participants’ 
reaction would have been to the placebo and to the drug. - It doesn’t account 
for the possibility that some individuals may have specific genes that may 
have enhanced or hindered an effect.

**Solution:** Best way to overcome this is by increasing the sample sizes
to ensure that more genetic diversity and range  is accounted for. It will 
also be good to have a similar experiment but with genetically identical mice.

Controlling for confounders - As additional information such as if participants’ 
parents were addicts, participant’s location, social environment and mental health 
were not collected leaving room for confounders. Other confounders include how long 
the person has used cannabis, long time users may be resistant to lower doses of THC 
(like long-time opioid users with decreasing sensitivity to opioids), making 
nabiximol ineffective in satisfying withdrawal symptoms and cravings. 


**Solution:** Collect associated information to control for confounders.

# References 
Weinstein, A. M., & Gorelick, D. A. (2011). Pharmacological Treatment of Cannabis 
Dependence. Current Pharmaceutical Design, 17(14), 1351–1358. 
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3171994/

Nicholas Lintzeris, M. B. B. S. (2019, September 1). Nabiximols for the treatment 
of Cannabis Dependence. JAMA Internal Medicine. Retrieved September 29, 2021, 
from https://jamanetwork.com/journals/jamainternalmedicine/fullarticle/2737918

The previous sources validate our findings in that they agree drug treatments 
which target THC successfully reduce withdrawal symptoms in cannabis-addicted 
individuals and increase the chances of their continued sobriety. 

# Acknowledgements
Our group met every Tuesday at 11am via zoom for one hour, having prepared 
beforehand to share ideas and progress made within the project. We all 
contributed to each aspect of the project but Nusrat focused on video editing 
and research, Lachy on script writing and presenting, Thomas on coding and 
visualisations, Scarlett on the RMarkdown and visualizations, and Lily on 
finalizing the project and presentation. 
